### 2024-11-27 parts/correction.tex
new proof

\AP This section is devoted to proving that polynomials in $\CorrectPoly$ can
be \kl{represented} by \kl{star-free $\Nat$-polyregular functions}. The key
lemma of this section is \cref{lem:correct-to-n-rat}, which is proved by
induction on the number of indeterminates of a given polynomial $P$. In order
to prove that result, we use the combinatorial
\cref{derivation-stabilises-correct:lem} that allows us to transform a
polynomial $P \in \CorrectPoly$ into a polynomial in $\Nat[\vec{X}]$ through a
well-chosen translation of the indeterminates. This argument is based on the
notion of \intro{discrete derivative} which is built by translating the domain
of the polynomial. To that end, let us write $\intro*\translate{K}$ for the
\intro{translation function} that maps a polynomial $P \in \Rel[X_1, \dots,
X_n]$ to the polynomial $P(X_1 + K, \dots, X_n + K)$.

@@

\begin{lemma}
	\label{all-positive-derivative:lem}
	Let $P \in \Nat[\vec{X}]$ that is non-constant, and $K \in \Nat$,
	then $\Diff{K}{P} \in \Nat[\vec{X}]$ and all of its
	coefficients are (positive) multiples of $K$.
	Furthermore, every monomial that strictly divides some monomial of $P$
	appears in $\Diff{K}{P}$.
\end{lemma}
\begin{proof}
	We prove the result for monomials, as it extends
	to $\Nat$-linear combinations by linearity.
	Let $P = \prod_{i = 1}^k X_i^{\alpha_i}$ be a monomial.
	Notice that $\translate{K}(P) = \prod_{i = 1}^n (X_i + K)^{\alpha_i}$,
	and using a binomial expansion
	we list all the possible divisors of $P$,
	all of which with coefficients that are positive integers and multiples of $K$ except the coefficient
	of the maximal monomial (equal to $P$ itself) which is $1$.
	As a consequence, $\translate{K}(P) - P$ is simply
	obtained by removing this maximal monomial, which concludes the proof.
\end{proof}

@@

	Let us assume that the alphabet over which the (\kl{commutative}) functions
	$f_\Delta$ and $f_\nu$ are defined is $\set{a_1, \dots, a_n}$, with $a_i$
	representing the indeterminate $X_i$ of the polynomials. Now, let us define
	by case analysis the following \kl{commutative} \kl{star-free
		$\Nat$-polyregular function}, defined on words $w$ of the form $w \defined
		a_1^{x_1} \cdots a_n^{x_n}$, with $x_1, \dots, x_n \geq 0$.

	\begin{equation*}
		f(w) \defined
		\begin{cases}
			f_{[X_i \mapsto x_i]}(w)                     & \text{ if } \exists i \in \set{1, \dots, n}, x_i \leq K \\
			f_\Delta(a_1^{x_1 - K} \cdots a_n^{x_n - K}) & \text{ otherwise }
		\end{cases}
		\quad .
	\end{equation*}
	Remark that
	$f$ is a \kl{commutative} \kl{star-free $\Nat$-polyregular function}
	that
	\kl{represents} $P$.
\end{proof}

>>>>

\AP This section is devoted to proving that polynomials in $\CorrectPoly$ can
be \kl{represented} by \kl{star-free $\Nat$-polyregular functions}. The key
lemma of this section is \cref{lem:correct-to-n-rat}, which is proved by
induction on the number of indeterminates of a given polynomial $P$. In order
to prove that result, we use the combinatorial
\cref{derivation-stabilises-correct:lem} that allows us to transform a
polynomial $P \in \CorrectPoly$ into a polynomial in $\Nat[\vec{X}]$ through a
well-chosen translation of the indeterminates. This argument is based on the
notion of \intro{discrete derivative} which is built by translating the domain
of the polynomial. To that end, let us write $\intro*\translate{K}$ for the
\intro{translation function} that maps a polynomial $P \in \Rel[X_1, \dots,
X_k]$ to the polynomial $P(X_1 + K, \dots, X_k + K)$.

@@

\begin{lemma}
	\label{all-positive-derivative:lem}
	Let $P \in \Nat[\vec{X}]$ that is non-constant, and $K \in \Nat$,
	then $\Diff{K}{P} \in \Nat[\vec{X}]$ and all of its
	coefficients are (positive) multiples of $K$.
	Furthermore, every monomial that strictly divides some monomial of $P$
	appears in $\Diff{K}{P}$.
\end{lemma}
\begin{proof}
	We prove the result for monomials, as it extends
	to $\Nat$-linear combinations by linearity.
	Let $P = \prod_{i = 1}^k X_i^{\alpha_i}$ be a monomial.
	Notice that $\translate{K}(P) = \prod_{i = 1}^k (X_i + K)^{\alpha_i}$,
	and using a binomial expansion
	we list all the possible divisors of $P$,
	all of which with coefficients that are positive integers and multiples of $K$ except the coefficient
	of the maximal monomial (equal to $P$ itself) which is $1$.
	As a consequence, $\translate{K}(P) - P$ is simply
	obtained by removing this maximal monomial, which concludes the proof.
\end{proof}

@@

	Let us assume that the alphabet over which the (\kl{commutative}) functions
	$f_\Delta$ and $f_\nu$ are defined is $\set{a_1, \dots, a_k}$, with $a_i$
	representing the indeterminate $X_i$ of the polynomials. Now, let us define
	by case analysis the following \kl{commutative} \kl{star-free
		$\Nat$-polyregular function}, defined on words $w$ of the form $w \defined
		a_1^{x_1} \cdots a_k^{x_k}$, with $x_1, \dots, x_k \geq 0$.

	\begin{equation*}
		f(w) \defined
		\begin{cases}
			f_{[X_i \mapsto x_i]}(w)                     & \text{ if } \exists i \in \set{1, \dots, k}, x_i \leq K \\
			f_\Delta(a_1^{x_1 - K} \cdots a_k^{x_k - K}) & \text{ otherwise }
		\end{cases}
		\quad .
	\end{equation*}
	Remark that
	$f$ is a \kl{commutative} \kl{star-free $\Nat$-polyregular function}
	that
	\kl{represents} $P$.
\end{proof}

### 2024-11-27 parts/starfree.tex
new proof

In this section, we leverage the decidability results of \cref{polynomials:sec}
to decide membership in $\NPoly$
inside $\ZPoly$ and membership in $\NSF$ inside $\NPoly$, both under the extra
assumption of \kl{commutativity}.

%\subsection{Deciding $\NPoly$ inside $\ZPoly$}

To characterize $\NPoly$ inside $\ZPoly$ we introduce the notion of
\kl{$(k,\Nat)$-combinatorial} function
(\cref{k-combinatorial:def}), following the spirit of previous
characterizations of subclasses of $\ZPoly$ in terms of \emph{polynomial
pumping arguments} \cite{DOUE21,DOUE22,CDTL23}.

@@

\AP
\Cref{k-combinatorial:def} is meaningful under the hypothesis of
\kl{commutativity}, because the latter reduces $\ZPoly$ to a combination of a
finite number of polynomials, arranged by counting modulo some numbers. This
characterization is stated in \cref{decompose-polynomial:lem}, and will
be the key ingredient in the decision procedures of $\NPoly$ inside $\ZPoly$.
To actually state the lemma, we will need some notations that will
formalize the intuitive idea of identifying tuples of integers modulo some
constant. Let $\omega \in \Nat_{\geq 1}$, we define the set
$\intro*\ModuloTypes[\omega] \defined \set{0, \dots, \omega^2} \times \set{0,
\dots, \omega - 1}$ of \intro{types modulo $\omega$}. To a number $n \in \Nat$,
we associate its \reintro{$\omega$-type} written
$\intro*\moduloType[\omega](n)$ which is defined as the pair $(t, r)$ where $t
= \min (n, \omega^2)$ and $r$ is the  remainder of the Euclidean division of
$n$ by $\omega$. This notion of type is lifted to vectors in $\Nat^p$ for any
$p \in \Nat$ by pointwise application.

\begin{lemma}
    \label{decompose-polynomial:lem}
    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Rel$-polyregular function},
    where we fix the alphabet $\Sigma = \set{a_1, \dots, a_n}$.
    There exists a computable
    $\omega \in \Nat_{\geq 1}$,
    and computable 
    \kl{integer binomial polynomials} $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^n}$
    such that for all $x_1, \dots, x_n \geq 0$:
    \begin{equation*}
        f\left(a_1^{x_1} \cdots a_n ^{x_n}\right) 
        = P_{\moduloType[\omega](x_1, \dots, x_n)}
        \left(
            \floor{x_1/\omega}, \dots, \floor{x_n/\omega}
        \right)
        \quad .
    \end{equation*}
\end{lemma}
\begin{proof}
    Let us first write $f \in \ZPoly$ as a linear combination 
    $f = f_+ - f_-$ where both $f_+$ and $f_-$ are in $\NPoly$.
    We 
    compute $\omega_+$ (resp. $\omega_-$)
    by
    applying
    \cref{n-poly-combinatorics:lem} 
    to $f_+$ (resp. to $f_-$).
    Finally, let $\omega$ be the least common multiple
    of $\omega_+$ and $\omega_-$, i.e. $\omega \defined \ppcm(\omega_+, \omega_-)$.
    Using this $\omega$,
    for any choice of
    $r \defined (r_1, \dots, r_n) \in \set{0, \dots, \omega - 1}$,
    there exists two \kl{natural binomial functions} $F_r^+$ and $F_r^-$
    associated to $f_+$ and $f_-$,
    such that $x_1, \dots, x_p \geq 1$:
    \begin{equation*}
        f\left(
        a_1^{\omega x_1 + r_1} \cdots a_n ^{\omega x_n + r_n}\right) 
        = F_r^+(x_1, \dots, x_p) - F_r^-(x_1, \dots, x_p) \quad .
    \end{equation*}
    Now, for large enough values of $x_1, \dots, x_p$,
    this means that 
    $f\left(
        a_1^{\omega x_1 + r_1} \cdots a_n ^{\omega x_n + r_n}\right)$
    coincides with an \kl{integer binomial polynomial},
    and we can therefore conclude selecting different values of $r$ and
    setting some subsets of variables to be assigned to the constant $0$.
\end{proof}

\begin{theorem}
    \label{decidable-n-poly:thm}
    Let $k \in \Nat$, and $f \in \ZPoly[k]$ be \kl{commutative}.
    Then, the following are equivalent:
    \begin{enumerate}
        \item \label{f-combinatorial:item} $f$ is \kl{$(k,\Nat)$-combinatorial},
        \item \label{f-npoly-combi:item} $f \in \NPoly[k]$,
    \end{enumerate}
    Furthermore, the properties are decidable,
    and conversions effective.
\end{theorem}
\begin{proof}
    The implication
    \cref{f-npoly-combi:item}
    $\implies$ \cref{f-combinatorial:item} immediately follows
    from 
    \cref{n-poly-combinatorics:lem} and \cref{decide-rat-poly-npoly:cor}.
    For the converse implication, we use \cref{decompose-polynomial:lem} 
    effectively compute a polynomial representation of $f$. That is, we obtain
    an $\omega_1 \in \Nat$ and \kl{integer binomial polynomials}
    $\seqof{P_{t}}{t \in \ModuloTypes[\omega_1]^k}$. Now, because $f$ is
    \kl{$(k,\Nat)$-combinatorial}, we also have a number $\omega_2 \in \Nat$
    such that for all \kl{pumping patterns} $q$, there exists a
    \kl{strongly natural binomial polynomial} $P$ such that $f \circ q(\omega_2 \vec{X})
    = P$ over $(\Nat_{\geq 1})^k$. By considering $\omega \defined
    \omega_1 \times \omega_2$, we can without loss of generality assume that
    $\omega_1 = \omega_2$ in the rest of this proof.

    Now, consider $t \in \ModuloTypes[\omega]^k$. By
    definition, this provides us with a set $S \subseteq \set{1, \dots, k}$,
    and numbers $r_1, \dots, r_k$ between $0$ and
    $\omega^2-1$ such that for all $x_1, \dots, x_k \geq 1$, the tuple $(x_1
    \omega \ind{S}(1) + r_1, \dots, x_k \omega \ind{S}(k) + r_k)$ has type $t$.
    In particular, for all $x_1, \dots, x_k \geq 1$, the following
    equality holds:
    \begin{equation*}
        f\left(\prod_{i = 1}^k 
            a_i^{ x_i \omega \ind{S}(i) + r_i}
        \right)
        =
        P_t(\seqof{x_i}{i \in S})
    \end{equation*}

    Let us therefore consider the \kl{pumping pattern} $q \colon \Nat^k \to
    \Sigma^*$ that is simply defined as $q(X_1, \dots, X_k) \defined \prod_{i =
    1}^k a_i^{ X_i \omega \ind{S}(i) + r_i}$. Because $f$ is
    \kl{$(k,\Nat)$-combinatorial} (with parameter $\omega$), there exists a
    \kl{strongly natural binomial polynomial} $P$ such that for all $x_1,
    \dots, x_n \geq 1$, the equality $f \circ q(\omega x_1, \dots, \omega x_k)
    = P(x_1, \dots, x_k)$ holds. In particular, we conclude that over
    $(\Nat_{\geq 1})^{\card{S}}$ the polynomial $P_t$ and $P$ coincide. They
    are therefore equal as multivariate polynomials, and we conclude that $P_t$
    is a \kl{strongly natural binomial polynomial} \kl{$\Nat$-rational
    polynomial}.

    Let us now construct a new way to compute $f$, by first
    computing the \kl{$\omega$-type} of the input (which is possible in
    $\MSO$), and then computing the suitable \kl{strongly natural binomial polynomials}
    which is possible in $\NPoly$
    thanks to \cref{decide-rat-poly-npoly:cor}.
\end{proof}

>>>>

In this section, we leverage the decidability results of \cref{polynomials:sec}
to decide membership in $\NPoly$
inside $\ZPoly$ and membership in $\NSF$ inside $\NPoly$, both under the extra
assumption of \kl{commutativity}. 
To characterize $\NPoly$ inside $\ZPoly$ we introduce the notion of
\kl{$(k,\Nat)$-combinatorial} function
(\cref{k-combinatorial:def}), following the spirit of previous
characterizations of subclasses of $\ZPoly$ in terms of \emph{polynomial
pumping arguments} \cite{DOUE21,DOUE22,CDTL23}.

@@

\AP Given a number $\omega \in \Nat$, let us write
$\intro*\ModuloTypes[\omega]^k$ for the collection of pairs $(S, \vec{r})$
where $S \subseteq \set{1, \dots, k}$ and $r \in \set{0, \dots, \omega - 1}^k$.
To a tuple $\vec{x} \in \Nat^k$, one can associate its \intro{$\omega$-type}
$\intro*\moduloType[\omega](\vec{x})$ which is the pair $(S, \vec{r})$ where $S
= \set{i \in \set{1, \dots, k} \mid x_i \geq \omega}$ and $\vec{r} = (x_i \mod
\omega)_{i \in \set{1, \dots, k}}$.

\begin{lemma}
    \label{decompose-polynomial:lem}
    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Rel$-polyregular function},
    where we fix the alphabet $\Sigma = \set{a_1, \dots, a_k}$.
    There exists a computable
    $\omega \in \Nat_{\geq 1}$,
    and computable 
    \kl{integer binomial polynomials} 
    $P_{(S,\vec{r})} \in \Rel[\seqof{X_i}{i \in S}]$ for $(S,\vec{r}) \in \ModuloTypes[\omega]^k$,
    such that for all $\vec{x} \in \Nat^k$,
    \begin{equation*}
        f\left(
            \prod_{i = 1}^k a_i^{x_i}
        \right)
        = P_{(S, \vec{r})}
        \left(
            \seqof{\floor{x_i / \omega}}{i \in S}
        \right)
        \text{ where } (S, \vec{r}) = \moduloType[\omega](\vec{x})
        \quad .
    \end{equation*}
\end{lemma}
\begin{proof}
    \textbf{TODO}. 
\end{proof}

\begin{theorem}
    \label{decidable-n-poly:thm}
    Let $k \in \Nat$, and $f \in \ZPoly[k]$ be \kl{commutative}.
    Then, the following are equivalent:
    \begin{enumerate}
        \item \label{f-combinatorial:item} $f$ is \kl{$(k,\Nat)$-combinatorial},
        \item \label{f-npoly-combi:item} $f \in \NPoly[k]$,
    \end{enumerate}
    Furthermore, the properties are decidable,
    and conversions effective.
\end{theorem}
\begin{proof}
    The implication
    \cref{f-npoly-combi:item}
    $\implies$ \cref{f-combinatorial:item} immediately follows
    from 
    \cref{n-poly-combinatorics:lem} and \cref{decide-rat-poly-npoly:cor}.
    For the converse implication, we use \cref{decompose-polynomial:lem} 
    effectively compute an $\omega \in \Nat$ and \kl{integer binomial polynomials}
    $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^k}$.

    Now, consider $(S, \vec{r}) \in \ModuloTypes[\omega]^k$. 
    Notice that for all vectors $\vec{x} \in (\Nat_{\geq 1})^k$, 
    the vector $(x_1 \omega \ind{S}(1) + r_1, \dots, x_k \omega \ind{S}(k) + r_k)$ has
    \kl{$\omega$-type} $(S, \vec{r})$.
    In particular, the following
    equality holds:
    \begin{equation*}
        f\left(\prod_{i = 1}^k 
            a_i^{ x_i \omega \ind{S}(i) + r_i}
        \right)
        =
        P_{(S, \vec{r})}(\seqof{x_i}{i \in S})
        \quad  \forall \vec{x} \in (\Nat_{\geq 1})^k
        \quad .
    \end{equation*}

    Let us therefore consider the \kl{pumping pattern} $q \colon \Nat^k \to
    \Sigma^*$ that is simply defined as $q(X_1, \dots, X_k) \defined \prod_{i =
    1}^k a_i^{ X_i \omega \ind{S}(i) + r_i}$. 
    Because $f$ is \kl{$(k,\Nat)$-combinatorial}, there exists an $\omega_2 \in
    \Nat$ and a \kl{strongly natural binomial polynomial} $P \in \Rel[X_1,
    \dots, X_k]$ such that $f \circ q = P$ over $(\omega_2 \Nat_{\geq 1})^k$.
    In particular, the following
    equality holds:
    \begin{equation*}
        P_{(S, \vec{r})}(\seqof{\omega_2 x_i}{i \in S})
        = 
        P(\omega_2 x_1, \dots, \omega_2 x_k)
        \quad  \forall \vec{x} \in (\Nat_{\geq 1})^k
        \quad .
    \end{equation*}
    Because $\omega_2 \neq 0$, we conclude that $P_{(S, \vec{r})} = P$
    and that both are polynomials over $\Rel[\seqof{X_i}{i \in S}]$,
    and furthermore that $P_{(S, \vec{r})}$
    is a \kl{strongly natural binomial polynomial}.

    To conclude, one can compute $f$ by first
    computing the \kl{$\omega$-type} of the input which is possible using a \kl{polyregular function},
    and then computing the suitable \kl{strongly natural binomial polynomials}
    which is possible in $\NPoly$
    thanks to \cref{decide-rat-poly-npoly:cor}. The resulting composition 
    belongs to $\NPoly$ thanks to
    \cref{stability-polyregular:lemma}.

    Let us now argue that the properties are decidable. Indeed, consider a
    function $f \in \ZPoly[k]$. Again, using
    \cref{decompose-polynomial:lem}, we can effectively compute the
    $\omega \in \Nat$ and the \kl{integer binomial polynomials}
    $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^k}$. If all of those polynomials
    are \kl{strongly natural binomial polynomials} (which is 
    decidable by \cref{decide-rat-poly-npoly:cor}),
    then $f \in \NPoly[k]$ as
    argued above. Conversely, if $f \in \NPoly[k]$, then $f$ is
    \kl{$(k,\Nat)$-combinatorial} by \cref{n-poly-combinatorics:lem}, and we
    argued that this implies that $P_{(S, \vec{r})}$ is a \kl{strongly natural
    binomial polynomial} for all $(S, \vec{r}) \in \ModuloTypes[\omega]^k$.
\end{proof}
